Retrofitting Effect Handlers onto OCaml

### 2020-11-17 retro-concurrency.tex
WIP trace in supplement

\crefformat{section}{§#2#1#3}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}

@@

\newcommand{\ostep}{\xrightarrow{o}}
\newcommand{\cstep}{\xrightarrow{c}}
\newcommand{\step}{\rightarrow}

@@

      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
    \end{array}
  \end{smathpar}
  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\end{minipage}
\caption{Operational semantics of Multicore OCaml effect handlers.}
\end{figure*}

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation |k''|.
Observe that the captured continuation |k''| includes the current handler.
Intuitively, the handler wraps around captured continuation. This gives
Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
captured continuation |k''| in the process. Recall that we do not capture C
frames as part of a continuation. To this end, \textsc{EffUnHn} models
unhandled effect. If the effect bubbles up to the top fiber, which we know does
not have an effect case (see \textsc{Callback} rule), we raise
\textsc{unhandled} exception at the point where the corresponding effect was
performed. This is achieved by appending the captured continuation to the front
of the current continuation.

@@

>>>>

\crefformat{section}{§#2#1#3}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

@@

\newcommand{\ostep}{\xrightarrow{o}}
\newcommand{\cstep}{\xrightarrow{c}}
\newcommand{\step}{\rightarrow}
\newcommand{\multistep}{\rightarrow\!\!*}

@@

      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{Unhandled}}{0}} \\
      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
    \end{array}
  \end{smathpar}
  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\end{minipage}
\caption{Operational semantics of Multicore OCaml effect handlers.}
\end{figure*}

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation |k''|.
Observe that the captured continuation |k''| includes the current handler.
Intuitively, the handler wraps around captured continuation. This gives
Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
captured continuation |k''| in the process. Recall that we do not capture C
frames as part of a continuation. To this end, \textsc{EffUnHn} models
unhandled effect. If the effect bubbles up to the top fiber, which we know does
not have an effect case (see \textsc{Callback} rule), we raise
\textsc{Unhandled} exception at the point where the corresponding effect was
performed. This is achieved by appending the captured continuation to the front
of the current continuation.

@@

\subsection*{Dynamic Semantics -- Examples}

\begin{definition}[Initial configuration]
	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
\end{definition}

Intuitively, the initial environment is empty and the initial stack is an empty
C stack.

\begin{definition}[Final configuration]
	Given a configuration $\config$, we say that the configuration is final
	(written $\textsf{final}(\config)$) if ~$\config =
	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
	$\env$.
\end{definition}

Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
the transitive reflexive closure of the top-level reduction relation $\step$.

\subsubsection*{Example 1}

Let

\begin{smathpar}
	\begin{array}{rcll}
		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
			&   &            & \mid \caseval{v}{v}})~0
	\end{array}
\end{smathpar}

then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
42$.

\begin{proof}
	$init(e) = \configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$

	By \textsc{StepC}, \textsc{AdminC}, \textsc{App1},

\begin{array}{rcll}
	configa{e}{\emptyset}{\cstacka{[]}{\bullet}} \\
	\step \\

\end{proof}
